---
title: "Getting Started with bbr.bayes and Stan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started-stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This "Getting Started with bbr.bayes" vignette takes you through some basic scenarios for modeling with Stan using `bbr.bayes`, introducing you to its standard workflow and functionality.

```{r setup, include=FALSE}
# NOTE: if running chunks interactively we need to load the package first
#   because renv isolation prevents us from finding an bbr.bayes installation
if (interactive() || (isTRUE(getOption('knitr.in.progress') && Sys.getenv("DRONE") != "true"))) {
  devtools::load_all()
}
```
```{r load packages, results = 'hide', message=FALSE, warning=FALSE}
library(bbr)
library(bbr.bayes)
library(dplyr)
library(purrr)
```

`bbr.bayes` is an extension to [`bbr`](https://metrumresearchgroup.github.io/bbr), which focuses on modeling using Bayesian methods. This vignette focuses on models defined in [Stan](https://mc-stan.org/). Note that you will also need to load `bbr` to use most `bbr.bayes` functionality.

A future `bbr.bayes` vignette will focus on modeling in NONMEM using Bayesian estimation methods. For details on modeling in NONMEM with non-Bayesian estimation methods, see the ["Getting Started with bbr and NONMEM" vignette](https://metrumresearchgroup.github.io/bbr/articles/getting-started.html).

# Setup

```{r set paths and define cleanup, include = FALSE}
MODEL_DIR <- system.file("model", "stan", package = "bbr.bayes")

# delete old files
cleanup <- function() {
  # delete new model
  if (fs::file_exists(file.path(MODEL_DIR, "mod1.yaml"))) fs::file_delete(file.path(MODEL_DIR, "mod1.yaml"))
  if (fs::dir_exists(file.path(MODEL_DIR, "mod1"))) fs::dir_delete(file.path(MODEL_DIR, "mod1"))

  # delete new copied model 
  if (fs::file_exists(file.path(MODEL_DIR, "fxaNew.yaml"))) fs::file_delete(file.path(MODEL_DIR, "fxaNew.yaml"))
  if (fs::dir_exists(file.path(MODEL_DIR, "fxaNew"))) fs::dir_delete(file.path(MODEL_DIR, "fxaNew"))

}

cleanup()
```


## Installing cmdstan

When modeling with Stan, `bbr.bayes` does _not_ currently use [`bbi`](https://metrumresearchgroup.github.io/bbr/articles/getting-started.html#installing-bbi), so you will not need it installed. However, you _will_ need `cmdstan` installed. Please see [the `cmdstanr` docs](http://mc-stan.org/cmdstanr/articles/cmdstanr.html#Installing_CmdStan) for instructions.

# Initial modeling run

## Create model object

To begin modeling, first create a model object using `new_model()`. This is an S3 object which you will pass to all of the other `bbr` and `bbr.bayes` functions that submit, summarize, or manipulate models.

The first argument (`.path`) must be the path to the directory containing the necessary files. For Stan models, you will also have to add `.model_type = "stan"` to the the `new_model()` call.

```{r first model}
mod1 <- new_model(file.path(MODEL_DIR, "mod1"), .model_type = "stan")
```

The call above will look for a directory called `{MODEL_DIR}/mod1`. _Inside_ this directory it will expect to find the files `mod1.stan` and `mod1-standata.R`. These "required files" will be discussed below, but you can find even more detail in [`?bbr_stan`](https://metrumresearchgroup.github.io/bbr.bayes/reference/bbr_stan.html). 

For now notice that, if any of them are _not_ found when you initialize the model, `new_model()` will create scaffolds (files with boilerplate non-working code) of the missing files, and print a message to the console telling you it has done so.

You can use the helpers described in [`?open_stan_file`](https://metrumresearchgroup.github.io/bbr.bayes/reference/open_stan_file.html) to open any of these scaffolded files for editing.

```{r, eval = FALSE}
open_standata_file(mod1)
```
```{r, echo = FALSE, comment=""}
mod1 %>%
  build_path_from_model("-standata.R") %>%
  readLines() %>%
  cat(sep = "\n")
```

The purpose of this scaffold is to provide some helpful framework to begin writing the necessary code for submitting a model with `bbr`. Below (in the `copy_model_from()` section) you will see how, once you have defined a model or two, you can use those models as your starting point for new models, instead of relying on writing new code from scaffolds.

### Starting from existing code

If you already have the relevant code for your new model, you can manually create these files in the output directory _before_ calling `new_model()`. In that case, you won't see any of the messages shown above and the `new_model()` call will simply create the `.yaml` file (described below) next to your directory. 

Alternatively, you can create the model with the scaffolds and use helpers like [`add_stanmod_file()`](https://metrumresearchgroup.github.io/bbr.bayes/reference/add_stanmod_file.html) to add your files to the model directory.

```{r, include = FALSE}
# clean up blank mod1 object and files
rm(mod1)
fs::file_delete(file.path(MODEL_DIR, "mod1.yaml"))
fs::dir_delete(file.path(MODEL_DIR, "mod1"))
```

### The `bbr` YAML file

This `new_model()` call will also create a `mod1.yaml` file that will sit next to this directory on disk. This YAML file  stores model metadata like description and tags (discussed below). For now the YAML file will look very bare, but hopefully its purpose will become more clear later.

## Reading in an existing model

Now assume that you have previously created a model and written all the necessary code. You can read this model into memory with `read_model()`.

```{r}
fxa <- read_model(file.path(MODEL_DIR, "fxa"))
```

If you take a look at _this model's_ `-standata.R` file, you can see an example of how this script might look in a real model.

```{r, eval = FALSE}
open_standata_file(fxa)
```
```{r, echo = FALSE, comment=""}
fxa %>%
  build_path_from_model("-standata.R") %>%
  readLines() %>%
  cat(sep = "\n")
```


## Submit model

For Stan models, model submission wraps the [cmdstanr::sample](https://mc-stan.org/cmdstanr/reference/model-method-sample.html) method. You can use `set_stanargs` to control which arguments are passed to `sample()`, and `get_stanargs()` to inspect the currently set arguments.

```{r}
fxa <- fxa %>%
  set_stanargs(list(seed = 123456,
                    iter_warmup = 100,
                    iter_sampling = 100))
```


### Using `check_stan_model()`

You can check whether all the required files are present at any time by calling the `check_stan_model()` function. By default this will also check the syntax of your `.stan` file, so that you can correct any errors before actually submitting the model.

```{r, eval = FALSE}
check_stan_model(fxa)
```

### Model submission

To submit a model to be run, you simply pass the model object to `bbr::submit_model()`.

```{r, eval = FALSE}
fxa_res <- fxa %>%
  submit_model()
```

```{r, echo = FALSE}
# load the res object
fxa_res <- read_fit_model(fxa)

# load the fake posteriors
fake_summary_df <- readRDS(system.file("vignette-fakes", "cmdstanr-summary-tibble.RDS",
                                       package = "bbr.bayes"))

# fake printout so that we don't have to actually run the model
system.file("vignette-fakes", "cmdstanr-sample-output.txt", package = "bbr.bayes") %>%
  readLines() %>%
  cat(sep = "\n")
```

As you can see, this also prints all of the diagnostic output from `$sample()` as it runs.

### Overwriting output from a previously run model

It is common to run a model, make some tweaks to it, and then run it again. However, to avoid accidentally deleting model outputs, `bbr` will error by default if it sees existing output when trying to submit a model. To automatically overwrite any previous model output, just pass `.overwrite = TRUE` to `submit_model()`.


## Summarize model

The `submit_model()` call will return the fit model object from `cmdstanr`. You can read all about this object in the [`CmdStanMCMC` object docs](https://mc-stan.org/cmdstanr/reference/CmdStanMCMC.html). It has several methods for diagnostics and model summaries.

### diagnostics from cmdstanr

```{r}
# this is just diagnostics
fxa_res$sampler_diagnostics() %>% str()
```


```{r, eval = FALSE}
# this pulls in the the actual posteriors
.s <- fxa_res$summary()
print(dim(.s))
```
```{r, echo = FALSE}
cat("[1] 7731   10")
```

```{r, eval = FALSE}
.s %>%
  slice_head(n = 15) %>%
  knitr::kable()
```
```{r, echo = FALSE, results = "asis"}
fake_summary_df %>%
  slice_head(n = 15) %>%
  knitr::kable()
```



### Using `read_fit_model()`

The `submit_model()` call automatically saves this fit object to disk as well. You can pass either a path or model object to `bbr.bayes::read_fit_model()` and it will re-load the `CmdStanMCMC` object into memory.

```{r}
new_res <- read_fit_model(file.path(MODEL_DIR, "fxa"))
new_res$sampler_diagnostics() %>% str()
```


### `model_summary()` function

We have **not yet implemented** `bbr::model_summary()` for `bbi_stan_model` objects. This is a work in progress, but until it is done, `model_summary()` currently returns the same fit object that was just discussed, and warns you to that effect.

```{r model_summary}
sum1 <- fxa %>% model_summary()
class(sum1)
```

```{r, eval = FALSE}
sum1$summary() %>%
  slice_head(n = 5) %>%
  knitr::kable()
```
```{r, echo = FALSE, results = "asis"}
fake_summary_df %>%
  slice_head(n = 5) %>%
  knitr::kable()
```

# Iteration

Much of the benefit of `bbr` is leveraged in the model iteration workflow, and the run log summaries that can be created afterwards.

## copy_model_from()

The `copy_model_from()` function creates a new model object based on a parent model, copying and renaming the necessary files in the process. The `copy_model_from()` call below will create an `fxaNew` directory and fill it as described in the `new_model()` section above. _However,_ instead of filling it with scaffolds, it will copy (and rename) all of the necessary files _from the parent model's directory_ into the new directory.

```{r copy_model 2}
fxaNew <- copy_model_from(
  .parent_mod = fxa,
  .new_model = "fxaNew",
  .inherit_tags = TRUE
)
```

> Note that, while the `.path` argument in  `new_model()` and `read_model()` is relative to your working directory, the `.new_model` argument of `copy_model_from()` is _relative to the directory containing the parent model._ This means that, assuming you would like to create the new model in the same directory as its parent, you only have to pass a filename (without extension) for the new model.

You can now open the relevant files and edit them with the desired changes, starting from where you left off with the parent model.

```{r, eval = FALSE}
open_stanmod_file(fxaNew)
```
```{r, echo = FALSE, message = FALSE}
# replace uniform with cauchy, for showing model_diff() below
mod_path <- build_path_from_model(fxaNew, ".stan")
mod_lines <- readLines(mod_path)
mod_lines %>%
  stringr::str_replace("uniform", "cauchy") %>%
  writeLines(mod_path)
```


### Comparing model files

Imagine that, after opening the `fxaNew.stan` file above, you decided to change the `emax` parameter from a uniform to a Cauchy distribution. The [`model_diff()`](https://metrumresearchgroup.github.io/bbr/reference/model_diff.html) function makes it easy to compare models to each other. For Stan models, `model_diff()` defaults to comparing the `<run>.stan` files.

```{r, results = "asis"}
model_diff(fxa, fxaNew)
```

You can also pass any of `c("standata", "init", "stanargs")` to have it compare the specified file instead.

```{r, results = "asis"}
# modify the iter_sampling arg and see the difference
set_stanargs(fxaNew, list(iter_sampling = 500))
model_diff(fxa, fxaNew, .file = "stanargs")
```

```{r}
# compare the data files and see that they are identical
model_diff(fxa, fxaNew, .file = "standata")
```

## Adding tags and notes

After looking at the results and diagnostics above, you can add tags, which can later be used to organize your modeling runs.

```{r add_tags 2}
fxa <- fxa %>% add_tags(c("fxaInhibit1Ncp", "unif emax"))
fxaNew <- fxaNew %>% add_tags(c("fxaInhibit1Ncp", "cauchy emax"))
```

Note that **using free text for your tags is discouraged**, for reasons mentioned in the [`?modify_tags` help page](https://metrumresearchgroup.github.io/bbr/reference/modify_tags.html). For simplicity's sake, we ignore that advice here, but please read it before using tags extensively in the wild. There is also a `tags_diff()` function for comparing 

In addition to tags, you can add notes that can be referenced later.

```{r}
fxaNew <- fxaNew %>%
  add_notes("Changed emax from uniform to cauchy distribution")
```

Importantly, **you must reassign to the model object** when calling any of these functions that modify it (notice the `fxaNew <- fxaNew ...` above). You can see a list of similar functions in the "See Also" section of [?modify_tags](https://metrumresearchgroup.github.io/bbr/reference/modify_tags.html) or [?modify_notes](https://metrumresearchgroup.github.io/bbr/reference/modify_notes.html).


### Inheriting tags

Note in the call above that we told `copy_model_from()` to inherit the tags from the parent model and automatically add them to the new model. You can also manually add tags (and notes) to this model.

```{r add_tags 3}
fxaNew <- fxaNew %>%
        add_tags("new gamma") %>%
        add_notes("Experimenting with different gamma distributions")

print(fxaNew$tags)
```

We can see that `fxaNew` contains the `"new gamma"` tag that we added manually _and_ the two tags that we inherited from its parent model. `copy_model_from()` also stores the model's "ancestry" which can be useful later in the project, as shown in the [Using the based_on field](https://metrumresearchgroup.github.io/bbr/articles/using-based-on.html) vignette.

# Run log

At any point, you can easily construct a "run log" tibble to summarize all models run up to this point.

Before we move on, note that you can get even more information about your models from the `config_log()` and `summary_log()` functions, as well as `add_config()` and `add_summary()` which automatically join the columns output from those functions against the tibble output from `run_log()`.  See the "Further Reading" section below for links to vignettes demonstrating those functions.

```{r run_log}
log_df <- run_log(MODEL_DIR)
log_df
```

The `run_log()` returns a tibble which can be manipulated like any other tibble. However, several of the columns (`tags`, `notes`, and `based_on` for example) are list columns, which complicates how you can interact with them. We provide some helper functions to more seamlessly interact with these log tibbles, as well as some sample `tidyverse` code below.

## Viewing tags example

The `bbr::collapse_to_string()` function can collapse any list column into a string representation of its contents. It is specifically designed for collapsing columns like `tags`, `notes`, and `based_on` into a more human-readable format. (We also use `knitr::kable()` here to render a prettier table in HTML, though this is _not_ necessary.)

```{r viewing tags new, results = "asis"}
log_df %>%
  collapse_to_string(tags, notes) %>%
  select(run, tags, notes) %>%
  knitr::kable()
```


## Filtering tags example

This code uses `purrr::map_lgl` to filter the run log to only rows containing a specific tag (`"unif_emax"`).

```{r filtering, results = "asis"}
log_df %>%
  filter(map_lgl(tags, ~ "unif emax" %in% .x)) %>%
  collapse_to_string(tags, notes) %>%
  select(run, tags, notes) %>%
  knitr::kable()
```


```{r cleanup bottom, include=FALSE}
# delete old files
cleanup()
```
