---
title: "Getting Started with bbr.bayes and Stan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bbr.bayes and Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette takes you through some basic scenarios for modeling with Stan using `bbr.bayes`, introducing you to its standard workflow and functionality.

```{r setup, include=FALSE}
# NOTE: if running chunks interactively we need to load the package first
#   because renv isolation prevents us from finding an bbr.bayes installation
if (interactive() || (isTRUE(getOption('knitr.in.progress') && Sys.getenv("DRONE") != "true"))) {
  devtools::load_all()
}
```
```{r load packages, results = 'hide', message=FALSE, warning=FALSE}
library(bbr)
library(bbr.bayes)
library(dplyr)
```

`bbr.bayes` is an extension to [`bbr`](https://metrumresearchgroup.github.io/bbr) which focuses on traceable and reproducible modeling using Bayesian methods. Note that you will also need to load `bbr` to use most `bbr.bayes` functionality.

This vignette focuses on models defined in [Stan](https://mc-stan.org/). 

A future `bbr.bayes` vignette will focus on modeling in NONMEM using Bayesian estimation methods. For details on modeling in NONMEM with non-Bayesian estimation methods, see the ["Getting Started with bbr and NONMEM"](https://metrumresearchgroup.github.io/bbr/articles/getting-started.html) vignette.

# Setup

```{r set paths and define cleanup, include = FALSE}
MODEL_DIR <- system.file("model", "stan", package = "bbr.bayes")

# delete old files
cleanup <- function() {
  # delete new model
  if (fs::file_exists(file.path(MODEL_DIR, "mod1.yaml"))) fs::file_delete(file.path(MODEL_DIR, "mod1.yaml"))
  if (fs::dir_exists(file.path(MODEL_DIR, "mod1"))) fs::dir_delete(file.path(MODEL_DIR, "mod1"))

  # delete new copied model 
  if (fs::file_exists(file.path(MODEL_DIR, "fxaNew.yaml"))) fs::file_delete(file.path(MODEL_DIR, "fxaNew.yaml"))
  if (fs::dir_exists(file.path(MODEL_DIR, "fxaNew"))) fs::dir_delete(file.path(MODEL_DIR, "fxaNew"))

  # reset args that we change in vignette
  try(
    set_stanargs(fxa, list(iter_warmup = 100,
                             iter_sampling = 100)), 
    silent = TRUE
  )
}

cleanup()
```


## Installing cmdstan

`bbr.bayes` uses cmdstan and the R package `cmdstanr` to run Stan models.  Please see [the cmdstanr docs](http://mc-stan.org/cmdstanr/articles/cmdstanr.html#Installing_CmdStan) for instructions on installing cmdstan.
(Note that, when modeling with Stan rather than NONMEM,
[`bbi`](https://metrumresearchgroup.github.io/bbr/articles/getting-started.html#installing-bbi)
is not required.)

# Creating a model object

The `bbr.bayes` workflow is organized around the concept of **model
objects**.  These are S3 objects which you will pass to `bbr` and
`bbr.bayes` functions that submit, summarize, or manipulate models.  A
model object points to a path on disk; the set of files that define a
model is derived from this path.  For a Stan model, the basic files
are:

 * `<run>.yml` - The `bbr` YAML file which stores model metadata like
   a description and tags.

 * `<run>/<run>.stan` - The Stan file.

 * `<run>/<run>-stanargs.R` - Contains a named list with all of the
   arguments that will be passed through to the [`$sample()`][sample]
   method of `?cmdstanr::CmdStanModel`.

 * `<run>/<run>-standata.R` - Contains all necessary R code to read in
   any source data and transform them to a Stan-ready data object
   (list).

 * `<run>/<run>-init.R` - This file contains all necessary R code to
   create the initial values passed to the cmdstanr's
   [`$sample()`][sample] method.

The [Stan models in `bbr` ](https://metrumresearchgroup.github.io/bbr.bayes/reference/bbr_stan.html) reference page provides detailed information about the files necessary for running a Stan model using `bbr`.


`bbr` model objects can either be created *de novo* or copied from an existing model object.  We outline the differences between these two approaches below.


## Creating a model object *de novo* 

To create a model object *de novo*, you use the `new_model()` function.

The `new_model()` function will look for (and, if not found, create) the files necessary to run a Stan model.  The first argument (`.path`) must be the path to the directory where you want to write these files.  For Stan models, you will also have to add `.model_type = "stan"` to the `new_model()` call.

```{r first model}
mod1 <- new_model(file.path(MODEL_DIR, "mod1"), .model_type = "stan")
```

The call above will look for a directory called `{MODEL_DIR}/mod1`. _Inside_ this directory it will expect to find the files `mod1.stan` and `mod1-standata.R`. These "required files" will be discussed below, and you can find even more detail in `?bbr_stan`. 

For now notice that, if any of them are _not_ found when you initialize the model, `new_model()` will create scaffolds (files with boilerplate, non-working code) of the missing files, and print a message to the console telling you it has done so.

You can use the helpers described in `?open_stan_file` to open any of these scaffolded files for editing.

```{r, eval = FALSE}
open_standata_file(mod1)
```

```{r, echo = FALSE, comment=""}
mod1 %>%
  build_path_from_model("-standata.R") %>%
  readLines() %>%
  cat(sep = "\n")
```

The purpose of this scaffold is to provide a helpful framework to begin writing the necessary code for submitting a model with `bbr.bayes`. Below (in the `copy_model_from()` section) you will see how, once you have defined a model, you can use that model as your starting point for new models, instead of relying on writing new code from scaffolds.


If you already have the relevant code for your new model, you can manually create these files in the output directory _before_ calling `new_model()`. In that case, you won't see any of the messages shown above and the `new_model()` call will simply create the `.yaml` file (described below) next to your directory.

Alternatively, you can create the model with the scaffolds and use helpers like `add_stanmod_file()` to add your files to the model directory. 

```{r, include = FALSE}
# clean up blank mod1 object and files
rm(mod1)
fs::file_delete(file.path(MODEL_DIR, "mod1.yaml"))
fs::dir_delete(file.path(MODEL_DIR, "mod1"))
```


## Reading in an existing model

Now assume that you have previously created a model and written all the necessary code. You can read this model into memory with `read_model()`.

```{r}
fxa <- read_model(file.path(MODEL_DIR, "fxa"))
```

If you take a look at _this model's_ `-standata.R` file, you can see an example of how this script might look in a real model.

```{r, eval = FALSE}
open_standata_file(fxa)
```

```{r, echo = FALSE, comment=""}
fxa %>%
  build_path_from_model("-standata.R") %>%
  readLines() %>%
  cat(sep = "\n")
```

## Creating a model by copy_model_from()

Much of the benefit of `bbr` is leveraged in the model iteration workflow, and the run log summaries that can be created afterwards.

The `copy_model_from()` function creates a new model object based on a parent model, copying and renaming the necessary files in the process. The `copy_model_from()` call below will create an `fxaNew` directory and fill it as described in the `new_model()` section above. _However,_ instead of filling it with scaffolds, it will copy (and rename) all of the necessary files _from the parent model's directory_ into the new directory.

```{r copy_model 2}
fxaNew <- copy_model_from(
  .parent_mod = fxa,
  .new_model = "fxaNew",
  .inherit_tags = TRUE
)
```

> Note that, while the `.path` argument in  `new_model()` and `read_model()` is relative to your working directory, the `.new_model` argument of `copy_model_from()` is _relative to the directory containing the parent model._ This means that, assuming you would like to create the new model in the same directory as its parent, you only have to pass a filename (without extension) for the new model.

You can now open the relevant files and edit them with the desired changes, starting from where you left off with the parent model.

```{r, eval = FALSE}
open_stanmod_file(fxaNew)
```

```{r, echo = FALSE, message = FALSE}
# replace uniform with cauchy, for showing model_diff() below
mod_path <- build_path_from_model(fxaNew, ".stan")
mod_lines <- readLines(mod_path)
mod_lines %>%
  stringr::str_replace("uniform", "cauchy") %>%
  writeLines(mod_path)
```


## The `bbr` YAML file

This `new_model()` and `copy_model_from()` calls will create the YAML file associated with the model object (e.g., `mod1.yaml`).  This YAML file  stores model metadata like the  description, notes, and tags (discussed below). The YAML file from a model created using `new_model()` will be very bare, while the `copy_model_from()` call will copy the YAML file from the parent model.  The purpose of the YAML file will become more clear when we begin comparing models and generating model run logs.


# Submitting a model

For Stan models, model submission wraps the [cmdstanr::sample()][sample] method.

### Passing arguments to `$sample()`

You can use `set_stanargs` to control which arguments are passed to `cmdstanr::sample()`.

```{r}
fxa <- fxa %>%
  set_stanargs(list(seed = 123456,
                    iter_warmup = 300,
                    iter_sampling = 300))
```

`bbr.bayes` stores these arguments with the model files, making it easy to reproduce your results later, by ensuring that you can run your model with the same arguments that were previously used. You can use `get_stanargs()` at any time to inspect the currently set arguments. 

```{r, results = 'asis'}
get_stanargs(fxa)
```


### Using `check_stan_model()`

You can check whether all the required files are present at any time by calling the `check_stan_model()` function. By default this will also check the syntax of your `.stan` file, so that you can correct any errors before actually submitting the model.

```{r, eval = FALSE}
check_stan_model(fxa)
```

### Model submission

To submit a model to be run, you simply pass the model object to [`submit_model()`].

```{r, eval = FALSE}
fxa_res <- fxa %>%
  submit_model()
```

```{r, echo = FALSE}
# load the res object
fxa_res <- read_fit_model(fxa)

# load the fake posteriors
fake_summary_df <- readRDS(system.file("vignette-fakes", "cmdstanr-summary-tibble.RDS",
                                       package = "bbr.bayes"))

# fake printout so that we don't have to actually run the model
system.file("vignette-fakes", "cmdstanr-sample-output.txt", package = "bbr.bayes") %>%
  readLines() %>%
  cat(sep = "\n")
```

As you can see, this also prints all of the diagnostic output from `$sample()` as it runs.

### Overwriting output from a previously run model

It is common to run a model, make some tweaks to it, and then run it again. However, to avoid accidentally deleting model outputs, `bbr` will error by default if it sees existing output when trying to submit a model. To automatically overwrite any previous model output, just pass `.overwrite = TRUE` to [`submit_model()`].


# Summarizing model output

The [`submit_model()`] call will return the fit model object from `cmdstanr`. You can read all about this object in the `?cmdstanr::CmdStanMCMC` object docs. It has several methods for diagnostics and model summaries.

### diagnostics from cmdstanr

```{r}
# this is just diagnostics
fxa_res$sampler_diagnostics() %>% str()
```


```{r, eval = FALSE}
# this pulls in the actual posteriors
.s <- fxa_res$summary()
print(dim(.s))
```
```{r, echo = FALSE}
cat("[1] 7731   10")
```

```{r, eval = FALSE}
.s %>%
  slice_head(n = 5) %>%
  knitr::kable()
```
```{r, echo = FALSE, results = "asis"}
fake_summary_df %>%
  slice_head(n = 15) %>%
  knitr::kable()
```

### `posterior::draws` methods

Many `{cmdstanr}` users will also be familiar with the `{posterior}` package and its [`draws` objects and methods](https://mc-stan.org/posterior/articles/posterior.html#draws-formats). `bbr.bayes` has also implemented the [`as_draws()`](https://metrumresearchgroup.github.io/bbr.bayes/reference/bbr_as_draws.html) method and its variants, which can be used on either a model object or the `?cmdstanr::CmdStanMCMC` object.

```{r, eval = FALSE}
d_array <- as_draws(fxa)
print(dim(d_array))
```
```{r, echo = FALSE}
cat("[1]  300   4   7731")
```

```{r, eval = FALSE}
d_df <- as_draws_df(fxa)
print(dim(d_df))
```
```{r, echo = FALSE}
cat("[1] 1200 7734")
```

### Using `read_fit_model()`

The [`submit_model()`] call automatically saves this fit object to disk as well. You can pass either a path or model object to `read_fit_model()` and it will re-load the `?cmdstanr::CmdStanMCMC` object into memory.

```{r}
new_res <- read_fit_model(file.path(MODEL_DIR, "fxa"))
new_res$sampler_diagnostics() %>% str()
```


### `model_summary()` function

We have **not yet implemented** `model_summary()` for `bbi_stan_model` objects. This is a work in progress, but until it is done, `model_summary()` currently returns the same fit object that was just discussed, and warns you to that effect.

```{r model_summary}
sum1 <- fxa %>% model_summary()
class(sum1)
```

```{r, eval = FALSE}
sum1$summary() %>%
  slice_head(n = 5) %>%
  knitr::kable()
```
```{r, echo = FALSE, results = "asis"}
fake_summary_df %>%
  slice_head(n = 5) %>%
  knitr::kable()
```


## Model annotation

Another useful feature of `bbr` is that it allows you to easily annotate and organize your models, for increased traceability and reproducibilty of your analysis. We demonstrate this a bit in this section, but it is described in more detail in both the ["Model Management" page of MeRGE Expo 1](https://merge.metrumrg.com/expo/expo1-nonmem-foce/posts/model-management-demo.html#annotating-your-model) and the ["Getting Started with `bbr`"](https://metrumresearchgroup.github.io/bbr/articles/getting-started.html#adding-tags-and-notes) vignette.

### Tags

Our first model already had two tags attached to it.

```{r add_tags 2}
fxa$tags
```

Note in the call above that we told `copy_model_from()` to inherit the tags from the parent model and automatically add them to the new model. You can use `replace_tag()` (or any of the other helpers in `?modify_tags`) to modify these at any time.

```{r add_tags 3}
fxaNew <- fxaNew %>%
        replace_tag("unif emax", "cauchy emax")

print(fxaNew$tags)
```

You can also use `tags_diff()` to quickly compare two models.

```{r}
tags_diff(fxa, fxaNew)
```

### Notes

In addition to tags, you can add notes that can be referenced later. While tags are intended to be structured and consistent, such that you can filter or organize models based on them (shown below), notes can be more "free form" narrative comments.

```{r}
fxaNew <- fxaNew %>%
  add_notes("Changed emax from uniform to cauchy distribution")
```
While both annotations can be easily accessed from the model object with `fxaNew$tags` or `fxaNew$notes`, notice that they are also displayed, along with other information, when you print the model object in your console

```{r}
fxaNew
```

### Aside on reassignment

Importantly, **you must reassign to the model object** when calling any of these functions that modify it (notice the `fxaNew <- fxaNew ...` above). You can see a list of similar functions in the "See Also" section of `?modify_tags` or `?modify_notes`.


# Run log

At any point, you can easily construct a "run log" tibble to summarize all models run up to this point. You can see much more detail about using run logs in both the ["Model Summary" page of MeRGE Expo 1](https://merge.metrumrg.com/expo/expo1-nonmem-foce/posts/model-summary.html#create-a-run-log) and the ["Getting Started with `bbr`"](https://metrumresearchgroup.github.io/bbr/articles/getting-started.html#run-log) vignette.

The `run_log()` returns a tibble which can be manipulated like any other tibble. You can also use the `.include` argument to filter to only specific tags or run names, and use formatting functions like `collapse_to_string()`, `knitr::kable()`, or many of the [functions in `pmtables`](https://merge.metrumrg.com/expo/expo1-nonmem-foce/posts/model-summary.html#creating-a-report-ready-run-log-table).

```{r viewing tags new, results = "asis"}
log_df <- run_log(MODEL_DIR, .include = c("fxa", "fxaNew"))

log_df %>%
  collapse_to_string(tags, notes) %>%
  select(run, tags, notes) %>%
  knitr::kable()
```


## Summary log

It is often useful to add some basic diagnostic information to a run log table. For Stan models, this can be done with `stan_summary_log()` or `stan_add_summary()`. In addition to some sampling diagnostics and metadata, this function will summarize the specified variables. 

```{r, eval = FALSE}
log_df <- log_df %>%
  stan_add_summary(variables = "emax")
```
```{r, echo = FALSE}
# load the fake summary table, because we don't actually have model outputs on disk
log_df <- readRDS(system.file("vignette-fakes", "stan-summary-tibble.RDS",
                                       package = "bbr.bayes"))
```
```{r, comment = ""}
names(log_df)
```

There are default functions used for the summarization, which are shown below. You can also pass your own summarization functions via the `summary_fns` argument. See `?stan_summary_log` for more details.

```{r}
log_df %>%
  select(run, num_divergent, starts_with("emax")) %>%
  knitr::kable()
```

## Comparing model files

Imagine that, after opening the `fxaNew.stan` file above, you decided to change the `emax` parameter from a uniform to a Cauchy distribution. The `model_diff()` function makes it easy to compare models to each other. For Stan models, `model_diff()` defaults to comparing the `<run>.stan` files.

```{r, results = "asis"}
model_diff(fxa, fxaNew)
```

You can also pass any of `c("standata", "init", "stanargs")` to have it compare the specified file instead.

```{r, results = "asis"}
# modify the iter_sampling arg and see the difference
set_stanargs(fxaNew, list(iter_sampling = 500))
model_diff(fxa, fxaNew, .file = "stanargs")
```

```{r}
# compare the data files and see that they are identical
model_diff(fxa, fxaNew, .file = "standata")
```



```{r cleanup bottom, include=FALSE}
# delete old files
cleanup()
```

[`submit_model()`]: https://metrumresearchgroup.github.io/bbr.bayes/reference/stan_submit_model.html
[sample]: https://mc-stan.org/cmdstanr/reference/model-method-sample.html
